Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-07 mr-review.Rmd
added liml text

The outcome variable can be binary, a count, or a continuous variable.

@@

Three assumptions are central to MR. Satisfying them is necessary to infer 
causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
The first assumption states that the exposure is associated with the outcome. The second requires 
independence, conditional on the exposure and all measured and unmeasured confounders, between the 
genetic marker and the outcome. The last of the three assumptions states that the genetic marker is 
independent of all (measured and unmeasured) confounders of the relationship between exposure and 
outcome. We next discuss these in more detail and consider their evaluation in MR studies [@vanderweele_methodological_2014].

If the first assumption, about the association between the genetic marker and the exposure, holds, 
but this association is weak, then an amplification of biases, such as those due to violations of 
assumptions 2 and 3, may result [@vanderweele_methodological_2014].

@@

@@

>>>>

The outcome variable can be binary, a count, or a continuous variable. For brevity and simplicity, 
we focus on continuous outcomes below.

@@

Three assumptions are central to MR. Satisfying them enables inference of 
causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
The first assumption states that the exposure is associated with the outcome. The second requires 
independence, conditional on the exposure and all measured and unmeasured confounders, between the 
genetic marker and the outcome. The last of the three assumptions states that the genetic marker is 
independent of all (measured and unmeasured) confounders of the relationship between exposure and 
outcome. We next discuss these in more detail and consider their evaluation in MR studies [@vanderweele_methodological_2014].

If the first assumption, about the association between the genetic marker and the exposure, holds, 
but this association is weak, then an amplification of biases, such as those due to violations of 
assumptions 2 and 3, may result [@vanderweele_methodological_2014]. @burgess_avoiding_2011, in analyses 
of simulated and real data, demonstrated that bias in causal estimates increases with decreases in 
the strength of association between the SNP and exposure.

@@

Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
estimates with their many desirable properties. Limited information maximum likelihood from econometrics 
is the earliest approach for likelihood-based inference in MR [@davidson1993estimation]. CoMM and PMR-Egger are two recent likelihood-based methods [@yang_comm_2019;@yuan_testing_2020].

Limited information maximum likelihood model equations are shown below (Equations \@ref(eq:lm5) and \@ref(eq:lm6)).

\begin{equation}
  x_i = \alpha_0 + \alpha_1 z_i + \epsilon_i
  (\#eq:lm5)
\end{equation}

\begin{equation}
  y_i = \beta_0 + \beta_1 \hat x_i + \tau_i
  (\#eq:lm6)
\end{equation}

They differ from the two-stage model because the two error terms, $\epsilon_i$ and $\tau_i$, have a bivariate normal
distribution.

CoMM, along with its extension for summary statistics, CoMM-S2, uses a collaborative mixed model to 
relate exposure and outcome variables to the SNP genotypes [@yang_comm-s2_2020]. Model fitting then 
proceeds via a parameter-expanded EM algorithm [@yang_comm_2019;@liu_parameter_1998].

We discuss PMR-Egger below in the context of methods that deal with horizontal pleiotropy.

@@

probabilistic Mendelian randomization (PMR-Egger) for two-sample MR with correlated SNPs models [@yuan_testing_2020]
